---
title: "Data Cleaning with dplyr (2)"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(haven)
library(dplyr)
library(ggplot2)
library(rsdmx)
knitr::opts_chunk$set(echo = FALSE)

# Loading conflict data from ICDP
temp <- tempfile() # opens empty temporary file
download.file("https://correlatesofwar.org/data-sets/national-material-capabilities/nmc-v5-1/at_download/file", temp) # downloads and saves into 'temp'
cow <- read.csv(unz(temp, # unzips & reads CSV file
                    "NMC_5_0.csv"),
                header = T)
unlink(temp)
rm(temp)

# Loading data from the CPDS
cpds <- read_dta("https://www.cpds-data.org/images/Update2020/CPDS_1960-2018_Update_2020.dta")

# Generating sample survey data for recode()
surv <- data.frame(idno = seq(1,12,1),
                   vaccinate = sample(rep(c("disagree strongly", "disagree", "agree", "agree strongly"),times = 3)))


```

## Introduction {data-progressive=TRUE}

In the previous tutorial, you learned how to do two important data cleaning operations with the `dplyr`-package: keeping (or dropping) variables using the `select()`-function, and keep or dropping observations based on some criteria using `filter()`. You also learned how you can efficiently execute several such commands in sequence by using the pipe-operator `%>%`.

With these commands, you can do some basic data cleaning and trim your dataset to what you really need for your analysis. But, obviously, this gets you only so far and most data cleaning and management also involves creating new variables, recoding existing ones, aggregating the data (e.g. into average values) as a whole or also for some groups of observations.

You will learn how to do these operations in this tutorial. And, as in the previous tutorial, we will rely mostly on functions from the `dplyr` package as well as the pipe operator `%>%`.

*A heads-up:* This tutorial is quite long and you might feel that it is more difficult compared to the previous ones. Make sure you go over this tutorial when you are well-rested and not in a hurry.

Ready to start?

## A quick recap of the `%>%` {data-progressive=TRUE}

This tutorial will make heavy use of the pipe operator `%>%`, so it is important to quickly repeat what it does. 

In essence, the pipe operator links together multiple functions in a row --- it tells R to take the result from one operation and feed it directly into the next one, and so on. By linking together multiple operations, we can build "data management pipelines", which take a messy dataset, clean it, and then spit out a clean dataset at the end.

To make matters more concrete, here is a piece of code that shows the pipe in action:
```{r pipeex, eval=F, echo=T}
cpds %>% # 1.
  select(country,year,unemp) %>%  # 2.
  filter(year>=1990) %>%  # 3.
  head() # 4.
```

In human language:

1. "Take the `cpds` object (the *Comparative Political Dataset*, in this case),..."
2. "...select three variables, `country`, `year`, and `unemp` (the unemployment rate), from it and drop the rest,..."
3. "...then keep only those observations that relate to 1990 and after,..."
4. "...and finally print the first few observations."

Notice that, because we use `%>%` operator, we can use the `head()` function by itself, without further details. R knows that we want to take the result from the previous steps here.

Got it? Great, then let's learn how to create new variables!

## Creating new variables with `mutate()` {data-progressive=TRUE}

For many data analysis projects, you need to construct your own variables or indicators from existing data. For example, you may need to divide one variable by another one to put them in relation to each other, or you may want construct an aggregate index to measure social and political attitudes such as authoritarianism (a desire for social order, conformity, and obedience to established norms and rules; [Feldman 2003](https://www.jstor.org/stable/3792510)) or ethnocentrism (the perception that one's own group is more intelligent, hard-working, and trustworthy than others; [Kam & Kinder 2012](https://doi.org/10.1111/j.1540-5907.2011.00564.x)). 

To create new variables, you can use the `mutate()` function from the `dplyr` package. As the name suggests, you use this function to "mutate" existing variables into new ones. Let's go straight to an example to show you how this works.

###

### Adding two variables

We start with a very simple operation: Calculating the sum of two variables. To practice this, we use the [*National Material Capabilities*](https://correlatesofwar.org/data-sets/national-material-capabilities) dataset from the *Correlates of War project*, which we also used in the previous tutorial.

Specifically, assume we wanted to add the following two variables from the dataset together:

* `milper`: The number of persons in countries' militaries (in thousands of persons)
* `upop`: The number of persons living in big cities (also in thousands of persons)

The code below shows how this is done (after some data cleaning, which you should now be familiar with):
```{r mutad-1, exercise=T}
cow %>% 
  select(stateabb,year,upop,milper) %>% 
  filter(stateabb=="USA" & year>=2005) %>% 
  mutate(added = milper + upop)
```

This code chunk should not be difficult to understand:

* The first lines select the variables we really need (and drop the rest) and filters the data so that we only have observations for the US for the years 2005 and after (`>=2005`)
* Then we create a new variable, `added`, that is the sum of `milper` and `upop`
* (We use the `%>%` to link everything elegantly together)

Easy!

###

Now over to you: Can you complete the code below to create a new variable, `summen`, which is the sum of:

* `irst`: iron & steel production, thousands of tons
* `pec`: primary energy consumption, tousands of coal-ton equivalents

```{r mutad_2, exercise=T}
cow %>% 
  select(stateabb,year,irst,pec) %>% 
  filter(stateabb=="USA" & year>=2005) %>% 
  mutate()
```

(Feel free to scroll up to see again how this is done.)

###

This is how you could have solved this:
```{r mutad_2a, eval=T,echo=T}
cow %>% 
  select(stateabb,year,irst,pec) %>% 
  filter(stateabb=="USA" & year>=2005) %>% 
  mutate(summen = irst + pec)
```



### Calculating a percentage

One very common type of "data mutation" you may have to do in your work is to calculate a percentage --- i.e. put two variables in relation to each other ("X in percent of Y"). This is especially important when we compare countries with larger and smaller economies or populations. 

We use again data from the *National Material Capabilities* dataset, specifically:

* `tpop`: The overall population, in thousands of persons
* `upop`: The urban population (persons living in big cities), in thousands of persons

If we wanted to compare how "urbanized" countries are, we would normally not look at the overall number but the *percentage* of persons living in big cities --- in this way, we can compare large countries such as the US or China to smaller countries such as Norway or Austria.

Here is how you could calculate the percentage with `mutate()`:
```{r mutex-1, exercise=T}
cow %>% 
  select(stateabb,year,tpop,upop) %>% 
  filter(stateabb=="NOR" & year>=2002) %>% 
  mutate(urban = (upop/tpop)*100)
```

Can you see what is going on here?

1. We first do some data cleaning, as before
2. Once we have trimmed the data, we use `mutate()` to create a new variable, `urban` as the percentage of the overall population that lives in larger cities (`upop` divided by `tpop`, then multiplied by 100)

From the result, you can see that about 17% of Norwegians lived in bigger cities in 2002 and that that number increased slowly over the following years.

Ready to try it for yourself?

###

We now use the *Comparative Political Data Set* (also familiar!). The dataset is already downloaded and saved as `cpds`. The code below already trims the data to some years for Norway and two central variables:

* `emp_ag`: The number of persons employed in agriculture, in thousands of persons
* `emp_civ`: The overall number of persons employed, in thousands of persons 

Can you complete the code below to calculate a new variable, `per_ag`, that measures the number of persons working in agriculture *as a percentage* of all employed persons?
```{r mutex-2, exercise=T}
cpds %>% 
  select(country,year,emp_ag,emp_civ) %>% 
  filter(country=="Norway" & year>=2010) %>% 
  mutate()
```

###

Here is one potential solution:
```{r mutex-2a, echo=T, eval=T}
cpds %>% 
  select(country,year,emp_ag,emp_civ) %>% 
  filter(country=="Norway" & year>=2010) %>% 
  mutate(per_ag = (emp_ag/emp_civ)*100)
```

### Other data transformations with `mutate()`

Next to calculating sums or percentages, you can also use the `mutate()` function for many other data transformations. See the documentation for an overview and some examples: https://dplyr.tidyverse.org/reference/mutate.html#useful-mutate-functions

## Summarizing with `summarize()` {data-progressive=TRUE}

As a data analyst, one of your central tasks is to aggregate and generalize, to turn a large and complex set of data into one or a few numbers that convey important information. For example, if you are working with a newly collected survey dataset, you may want to learn a bit about your sample as a whole: What is the average age of the respondents, what is the median income, how many respondents do you have in total? Similarly, if you work with cross-country comparative data, you may want to learn about the sample of countries you are working with: What is the average unemployment rate or what is the minimum and maximum observed level of spending on social protection?

The `summarize()` function allows you to do that, using the standard `R` functions:

* `mean()` to calculate the average
* `max()` to calculate the maximum
* `min()` to calculate the minimum
* `sd()` to calculate the standard deviation
* `median()` to calculate the median
* `n()` to count the number of observations

For example, you use `summarize(mean(unemployment))` to calculate the average of the `unemployment` variable, or `summarize(max(age))` to get the maximum of `age`.

Ready to try it out on some data?

###

### Calculate the average unemployment rate during the Great Recession

To practice using `summarize()`, let's use the Comparative Political Data Set to calculate the average unemployment rate in the advanced western democracies during the Great Recession --- the big *once-in-a-century* economic crisis (or so everyone thought...) following the collapse of the Lehman Bros. bank in 2008.

First, let us take a quick look at the data we will be working with:
```{r sum_dat, echo=F, eval=T}
cpds %>% 
  select(country,year,unemp,realgdpgr) %>% 
  filter(year==2009)
```

We will be using only data for the year 2009 (the year after the initial stock market crash) and two variables:

* `unemp`: The unemployment rate
* `realgdpgr`: real GDP growth (the economic growth rate)

You can flip through the table to see which countries we have in our data.

###

Now that you are familiar with the data, let's find out what `summarize()` does. Run the code to see for yourself:
```{r sumex-1, exercise = T}
cpds %>% 
  select(country,year,unemp) %>% 
  filter(year==2009) %>%  # we use only data from 
  summarize(mean(unemp))
```

You see that `summarize()` does as its name suggests: It aggregates a set of data down to a single number. In the previous example, this number was the aggregate unemployment rate in 2009.

### Creating new variables with `summarize()`

You can also save the result to a new variable. The process is the same as with `mutate()`. For example, to create a new variable `avg_unemp` containing the average of `unemp`, we would use: `summarize(avg_unemp = mean(unemp))`.

Again over to you: Can you complete the code below to calculate the average of the `realgdpgr` variable (which measures the rate of economic growth, or GDP growth) and save the result as `avg_growth`?
```{r sumex-2, exercise = T}
cpds %>% 
  select(country,year,realgdpgr) %>% 
  filter(year==2009) %>%  # we use only data from the year after the initial stock market crash
  summarize()
```

###

This is how you could have solved the previous exercise:
```{r sumex-2a, eval=T,echo=T}
cpds %>% 
  select(country,year,realgdpgr) %>% 
  filter(year==2009) %>%  # we use only data from the year after the initial stock market crash
  summarize(avg_growth = mean(realgdpgr))
```

You can see that 2009 was a serious crisis year: the advanced economies shrank by on average 4 percent!

### More than one summary value

You can also use `summarize()` to calculate two aggregate values at the same time. To do so, you just add a second calculation within the `summarize()` function. For example, the following code would calculate both the average and maximum unemployment rate:
```{r sumex-3, eval=F, echo=T}
summarize(avg_unemp = mean(unemp),
          max_unemp = max(unemp))
```

Give it a try: Complete the code below to calculate the maximum (`max()`) and minimum (`min()`) economic growth rate (`realgdpgr`) in the advanced economies in 2009:
```{r sumex-4, exercise=T}
cpds %>% 
  select(country,year,realgdpgr) %>% 
  filter(year==2009) %>%  # we use only data from the year after the initial stock market crash
  summarize()
```

###

This is one potential solution:
```{r sumex-4a, eval=T,echo=T}
cpds %>% 
  select(country,year,realgdpgr) %>% 
  filter(year==2009) %>%  # we use only data from the year after the initial stock market crash
  summarize(max_growth = max(realgdpgr),
            min_growth = min(realgdpgr))
```

You can see that there is quite some variation in our data! One country's economy shrank by almost 15 percent, while another country's economy even *grew* by almost three percent! (Of course, in this particular case, we would now also like to know which countries these are. But if we were working with anonymous survey data, knowing only the maximum and minimum value would be sufficient.) 


## Summarize by: `summarize()` & `group_by()` {data-progressive=TRUE}

Being able to summarize an entire dataset in one or two statistics is nice, but it is often not exactly what we want to do. Often, we want to calculate summary statistics for several groups of observations in our data. For example, if we are interested in the 'gender pay gap', we need to look at average salaries broken down *by gender*. Or, if we are working with cross-country data, we may want to calculate summary statistics for certain groups of countries.

Using `group_by()` with `summarize()` is relatively simple: We simply tell `R` that we want our summary statistics grouped by adding a `group_by()` command before `summarize()` in our "pipeline". For example:
```{r grex1, echo=T, eval=F}
group_by(gender) %>% 
  summarize(avg_salary = mean(salary))
```

Let's again work on a concrete example.

###

### The Great Recession in East and West

We work again with the CPDS dataset and the same main variables as before (`realgdpgr` & `unemp`), but we now add a third one: `poco`, which indicates if a country is *post-communist* or not. The variable is a so-called *dummy* variable: It takes the value of 1 if the country is post-communist, and 0 otherwise.

Take a look at the table to see how the data look like:
```{r grodat, eval=T, echo=F}
cpds %>% 
  select(country,year,unemp,realgdpgr,poco) %>% 
  filter(year==2009)
```

You see that the former Warsaw Pact countries in Central and Eastern Europe --- Poland, Lithuania, Bulgaria, etc. --- are labelled *post-communist*. Countries like Norway or France, of course, are not.

###

Now let us find out if the Great Recession hit harder in the formerly communist countries of Central and Eastern Europe.

The code below does what we covered in the previous section: It calculates the average unemployment rate across all countries in the CPDS dataset. 

Now, can you add code to get the result grouped by `poco`? (Hint: Add `group_b()` before `summarize()`; see also above)
```{r groex1, exercise = T}
cpds %>% 
  select(country,year,unemp,poco) %>% 
  filter(year==2009) %>% 
  summarize(avg_unem = mean(unemp))
```

###

Here is how you could have solved this one:
```{r groex1a, eval=T,echo=T}
cpds %>% 
  select(country,year,unemp,poco) %>% 
  filter(year==2009) %>% 
  group_by(poco) %>% # Add the group_by() command here
  summarize(avg_unem = mean(unemp))
```

You simply add a `group_by()` command with the relevant grouping variable into your "pipeline" before you summarize.

###

Let's do one more! Can you add code below to calculate the average GPD growth rate for post-communist countries and others? (Hint: Now you need to use both `group_by()` and `summarize()` by yourself!)
```{r groex2, exercise = T}
cpds %>% 
  select(country,year,realgdpgr,poco) %>% 
  filter(year==2009) %>% 
  
```

### 

Here is a possible solution:
```{r groex2a, eval=T,echo=T}
cpds %>% 
  select(country,year,realgdpgr,poco) %>% 
  filter(year==2009) %>% 
  group_by(poco) %>% 
  summarize(avg_growth = mean(realgdpgr))
```

Now you had to write a bit more code all by yourself. But this should still not have been too hard, or was it?

###

One last note on the `group_by()` function: It usually makes only sense to use a categorical (or perhaps ordinal) variable as a grouping variable --- a variable like post-communist, or also gender or education. *Technically*, you could also group by a metric/continuous variable (e.g., `unemp`). `R` will then calculate a summary statistic for each unique value of the continuous variable --- but this rarely makes sense, because continuous variables have, by definition, an infinite number of possible values. So, all you get is a long list of meaningless numbers.

###

Good job --- now you know create new variables (with `mutate()`), how to calculate summary statistics (with `summarize()`), and how to group the latter (with `group_by()`). You are getting close to being able to tackle your own data analysis project!

But there is one more important command you should learn: `recode()`. We will cover this in the last part of this tutorial.

## Recoding with `recode()` {data-progressive=TRUE}

Recoding is clearly one of the most common task in a data cleaning and analysis operation. When you recode a variable, you change its values according to some rule. For example, you sometimes need to recode a Likert-scale variable ("disagree strongly", "disagree", "agree", "agree strongly") into a simpler binary *dummy* version of itself ("against", "in favor") to do some analyses.

To recode variables with `dplyr`, we use --- you guessed it! --- the `recode()` function. It works on a fairly simple logic: You need to specify the variable you want to recode and one or more rules that tell `R` how different values should be changed. For example: `recoded_var <- recode(original_var,old_value = "new_value")`

You can in principle use `recode()` by itself when you work with a single vector ("variable"), outside of a dataset. Usually, however, we work within a dataset and use `recode()` to create new variables in our dataset. And, as you learned earlier: We use `mutate()` to create new variables in a dataset. This means that you combine `mutate()` and `recode()` when you recode variables in a dataset.

This will become clearer when we look at an example.

### 

###

To have an easier start, we work first with a small simulated dataset instead of a real one. Think of this as a small survey dataset called `surv` with twelve respondents and two variables:

* `idno`: The respondent's individual survey ID number
* `vaccinate`: Whether the respondent would get vaccinated against COVID-19 ("disagree strongly", "disagree", "agree", "agree strongly")

Here is how the data look like:
```{r recdat, eval=T, echo=F}
surv

```

###

Assume that we are feeling silly today and we want to recode the `vaccinate` variable so that every instance of "disagree" is replaced with "banana". 

The code below generates a new variable, `vacc_banana`, which is a recoded version of `vaccinate`. Run it and take a look at the result:
```{r recex1, exercise=T}
surv %>% 
  mutate(vacc_banana = recode(vaccinate,
         "disagree" = "banana"))
```

###

You should see a bit clearer how `recode()` works: We define a rule (`"disagree" = "banana"`) and `R` then goes over the variable we want to recode and uses our rule to change the values. And when we do not specify a rule for a particular value, `R` leaves that value alone.

Let's work on a more realistic example: Let's say we really want to simplify the `vaccinate` variable and reduce the number of categories. Specifically, we want to:

* Combine "disagree strongly" and "disagree" into "against"
* Combine "agree strongly" and "agree" into "in favor"

The code below shows you the first part --- can you complete the recoding rules yourself?
```{r recex2, exercise = T}
surv %>% 
  mutate(vacc_simple = recode(vaccinate,
                              "disagree strongly" = "against",
                              "disagree" = "against",
                              "agree" = 
                              ))
```

###

This is how you could have solved the exercise:
```{r recex2a, eval=T,echo=T}
surv %>% 
  mutate(vacc_simple = recode(vaccinate,
                              "disagree strongly" = "against",
                              "disagree" = "against",
                              "agree" = "in favor",
                              "agree strongly" = "in favor"))
```

Are you getting the hang of it? 

###

We can also use `recode()` to change numbers ("continuous" or "numeric" variables). In this case, we only need to place the old value within single quotation marks. The code below shows how you can change the value of 7 to 9999:
```{r recex3, exercise = T}
surv %>% 
  mutate(id_new = recode(idno,
                         `7` = 9999))
```

### Recoding `factor` variables 

Do you remember what factors were in `R`? Factors are how `R` likes to store categorical or ordinal variables such as a survey respondent's gender or their level of education. Factors are, in essence, numbers with labels.

Let's take a quick look at how factors work, to freshen up your memory. We will use the `vaccinate` variable from the earlier part for this.

The `vaccinate` variable was so far saved as a character ("text") vector. Let's see what it looks like by printing out the *unique* values of the `vaccinate` variable:
```{r fac1, echo=F, eval=T}
unique(surv$vaccinate)
```
You see that it has four unique levels: "disagree strongly", "disagree", "agree", "agree strongly".

If we would convert it to a factor, it would look like this:
```{r fac2, echo=F, eval=T}
vacc <- c("disagree strongly", "disagree", "agree", "agree strongly")

vacc_fac <- factor(vacc,
                   levels = vacc)


print(data.frame(levels = unique(vacc_fac), 
           value = as.numeric(unique(vacc_fac))))
```

Now we have associated the labels ("disagree strongly", "disagree", etc.) with numbers (1, 2, etc.). Again, factors are numbers with labels on them.

###

Recoding a factor works in principle like recoding other types of variables, and we can still use `recode()`. We just need to be more explicit about the recoding rules. Specifically, we need to tell `R` that it should leave all levels that we do not want to change as they are. For example:
```{r fac3, echo=T,eval=F}
recoded_factor <- recode(existing_factor,
                         "old_level" = "new_level",
                         .default = levels(existing_factor))
```

Now comes a bigger exercise:

1. The code below "mutates" the `vaccinate` variable into a factor called `vacc_fac`
2. Can you add code (using the example above) to create a "mutated" version of the new `vacc_fac` variable, where "disagree" is changed to to "banana"

```{r facex1, exercise=T}
surv %>% 
  select(vaccinate) %>% 
  mutate(vacc_fac = factor(vaccinate,
                           levels = c("disagree strongly", "disagree", "agree", "agree strongly"))) %>% 
  mutate(vacc_banana = recode(     ))
```

This one may be a bit trickier than the previous exercises. Take your time! (And don't get frustrated if you get error messages.)

###

Here is a possible way to solve it:
```{r facex1a, eval=T, echo=T}
surv %>% 
  select(vaccinate) %>% 
  mutate(vacc_fac = factor(vaccinate,
                           levels = c("disagree strongly", "disagree", "agree", "agree strongly"))) %>% 
  mutate(vacc_banana = recode(vacc_fac,
                              "disagree" = "banana",
                              .default = levels(vacc_fac)))
```

You see that the result is the same as earlier --- every instance of "disagree" is now labelled "banana". Only the type of variable is now a `fct`. 

To show you what is happening "under the hood", here are the new levels and underlying numbers associated with the levels of our new `vacc_fac` variable:
```{r facex1b, echo=F, eval=T}
vacc <- c("disagree strongly", "banana", "agree", "agree strongly")

vacc_fac <- factor(vacc,
                   levels = vacc)


print(data.frame(levels = unique(vacc_fac), 
           value = as.numeric(unique(vacc_fac))))
```

You see that the level "banana" is now associated with the number 2.

Factors are certainly a bit tricky to deal with --- but we do use them, and so it is important that you understand how to manage and recode them. Also, this exercise should have served as a refresher on factors in R.

Now, to engrave the lessons of this (admittedly) very long and not always easy tutorial: Debugging exercises!

## Debugging exercises {data-progressive=TRUE}

You should know the drill at this point: You get a few code chunks that do not work and are asked to correct them.

Ready?

###

### Problem No. 1

With this first chunk of code, we try to produce a "data cleaning pipeline" --- but it seems there is something missing at some point. Can you help?
```{r debug1, exercise=T}
cow %>% 
  select(stateabb,year,irst,pec)
  filter(stateabb=="USA" & year>=2005)
  mutate(summen = irst + pec)

```

###

Here is the solution: You needed to add pipes (`%>%`) at the of the middle two lines:
```{r debug1a, eval=T,echo=T}
cow %>% 
  select(stateabb,year,irst,pec) %>% # added
  filter(stateabb=="USA" & year>=2005) %>% # added
  mutate(summen = irst + pec)
```

### Problem No. 2

Now we want to calculate a percentage using `mutate()` --- but `R` does not seem happy. Can you fix this bug?
```{r debug2, exercise=T}
cpds %>% 
  select(country,year,emp_civ,emp_serv) %>% 
  filter(country == "Norway" & year>=2015) %>% 
  mutate(per_serv = (emp_serv/emp_civ*100)
```

###

This one might have difficult to spot --- there was a missing closing parenthesis `)`!
```{r debug2a, eval=T, echo=T}
cpds %>% 
  select(country,year,emp_civ,emp_serv) %>% 
  filter(country == "Norway" & year>=2015) %>% 
  mutate(per_serv = (emp_serv/emp_civ)*100)
```

Compare:
```{r debug2b, eval=F, echo=T}
mutate(per_serv = (emp_serv/emp_civ*100) # Wrong

mutate(per_serv = (emp_serv/emp_civ)*100) # Correct
```

### Problem No. 3

This one looks quite messy. Can you correct all the typos here?

```{r debug3, exercise=T}
cow %>% 
  seet(stateabb,year,irst,pec) %>% 
  fltry(stateabb=="NOR" & year==2007) %>% 
  mutte(summen = irst + pec)
```

###

No one likes proofreading --- but there was no other option here. All the commands were misspelled: 

* `select()`, not `seet()`
* `filter()`, not `fltry()`
* `mutate()`, not `mutte()`

```{r debug3a, eval=T, echo=T}
cow %>% 
  select(stateabb,year,irst,pec) %>% 
  filter(stateabb=="NOR" & year==2007) %>% 
  mutate(summen = irst + pec)
```

### Problem No. 4

One more problem with `mutate()` --- help please!

```{r debug4, exercise=T}
cpds %>% 
  select(country,year,ud,unemp) %>% 
  filter(country = "Norway" & year=2015) %>% 
  mutate(added = ud + unemp)
```

###

This one might have been a bit misleading. There was nothing wrong with `summarize()` --- the issue was in the `filter()`-command: We need to use double equal signs (`==`) here.
```{r debug4a, eval=T,echo=T}
cpds %>% 
  select(country,year,ud,unemp) %>% 
  filter(country == "Norway" & year==2015) %>% 
  mutate(added = ud + unemp)
```

### Problem No. 5

This is the last `mutate()` problem, promise!
```{r debug5, exercise=T}
cow %>% 
  select(stateabb,year,irst,pec) %>% 
  filter(stateabb=="NOR" & year==2007) %>% 
  mutate(summen = irst + stateabb)
```

###

The code as such was not the problem this time --- it was the operation within `mutate()` that made no sense. We tried to add the `irst` variable to the `stateabb` variable. In effect, we asked `R` to calculate `708 + NOR`. Obviously, we can only add numbers! Here is a calculation that functions: 

```{r debug5a, echo=T, eval=T}
cow %>% 
  select(stateabb,year,irst,pec) %>% 
  filter(stateabb=="NOR" & year==2007) %>% 
  mutate(summen = irst + pec)
```

### Problem No. 6

Now we want to summarize --- specifically, we want to get the median level of iron & steel production in all the countries in our *Material Capabilities* dataset (in a single year, 2002). We have some code, but it does not give us the desired result, just a long list of numbers.
```{r debug6, exercise = T}
cow %>% 
  select(stateabb,year,irst) %>% 
  filter(year==2002) %>% 
  summarize(irst)
```

###

Not too difficult, was it? If we want to use `summarize()` to get some summary statistics, we need to also specify *which* statistic we want: The average, the median, the sum, the minimum, maximum,...

Here is how we could have calculated the median:
```{r debug6a, eval=T, echo=T}
cow %>% 
  select(stateabb,year,irst) %>% 
  filter(year==2002) %>% 
  summarize(median(irst))
```

### Problem No. 7

Now we want to get the average number of military personnell (`milper`) in our dataset (again, for a single year). This time, we specified a summary function, the average. But `R` does not seem to like this....
```{r debug7, exercise=T}
cow %>% 
  select(stateabb,year,milper) %>% 
  filter(year==2002) %>% 
  summarize(average(milper))
```

###

In `R`, the function to calculate the average --- or mean --- is called `mean()`. Seems like a case of *tomayto, tomahto*, but this is just how `R` works. It wants the commands it knows, exactly the commands it knows, and nothing but the commands it knows.
```{r debug7a, eval=T, echo=T}
cow %>% 
  select(stateabb,year,milper) %>% 
  filter(year==2002) %>% 
  summarize(mean(milper))
```

### Problem No. 8

Now we go back to our simulated survey dataset, `surv`, and do some recoding. Here is the `surv` dataset, in case you forgot what it looks like:
```{r debug8d, eval=T, echo=F}
surv
```

To the problem: We are again feeling silly and want to recode the levels of our `vaccinate` variable as different types of fruit. But `R` does not play along. Nothing gets changed. 
```{r debug8, exercise=T}
surv %>% 
  mutate(vacc_simple = recode(vaccinate,
                              "apple" = "disagree strongly",
                              "banana" = "disagree",
                              "orange" = "agree",
                              "cherry" = "agree strongly"))
```

(This one might take a while to figure out, and you have to change quite a bit of code. Hang in there!)

###

Figured it out? It was the order of old and new levels: old comes before new! 

* *Wrong*: `"apple" = "disagree strongly"`
* *Correct*: `"disagree strongly" = "apple"`

Here is the entire chunk, corrected:
```{r debug8a, eval=T, echo=T}
surv %>% 
  mutate(vacc_simple = recode(vaccinate,
                              "disagree strongly" = "apple",
                              "disagree" = "banana",
                              "agree" = "orange",
                              "agree strongly" = "cherry"))
```

###

Congrats, you almost made it to the end of this tutorial! The last item on the list: Proper exercises, where you write your own code from scratch. 

But you might want to take a quick break and grab a cup of coffee first --- this was a long and certainly not always easy tutorial!

## Exercises {data-progressive=TRUE}

### Exercise 1
