---
title: "Visualizing data with ggplot2"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)

fruit <- data.frame(type = c("Apples","Oranges"),
                    number = c(42,17))

cpds <- haven::read_dta("https://www.cpds-data.org/images/Update2020/CPDS_1960-2018_Update_2020.dta") %>% 
  filter(year==2015 & poco==0) %>% 
  select(country,iso,year,womenpar,childcare_pmp,emprot_reg,unemp) %>% 
  rename(childcare = childcare_pmp)

```


# Visualizing data with ggplot2

## First steps {data-progressive=TRUE}

###

In this tutorial, you will learn the basics of visualizing data with `ggplot2`, which is one of the various packages for R, and is probably the most popular tool to make graphs in R with.

This tutorial for `ggplot2` is *interactive*, which means that you get to do many things yourself --- with help, of course! Throughout the tutorial, you find little consoles like the one just below.
```{r tutintro, exercise=T}

```

Each of these consoles is the same as the console on the bottom left in RStudio: It allows you to run small pieces of R code and get a result. 

Try it out: Type `1+1` into the console and click `Run`. Did you get a result? If yes, try also computing `23*89`. If you like, play around with this a bit. 

Otherwise, click 'Continue' to begin the tutorial.

### First steps: A simple bar graph

In the first part of this tutorial, you will learn how to create a simple bar graph, step-by-step. To show you what the end result is going to look like: This below is the graph that we will be creating.

```{r barshow, eval=T, echo=F}
ggplot(data = fruit, aes(x=type,y=number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```

This graph is obviously not very complicated: It shows how many pieces of two types of fruit (apples and oranges) we have.

Let's also take a quick look at the "dataset" that we use here:
```{r fruittab, echo=F,eval=T}

print.data.frame(data.frame(type = c("Apples","Oranges"),
                    number = c(42,17)))
```

Our "dataset" contains only two variables (`type` & `number`) and only two observations, one for apples and one for oranges. Basically, our dataset tells us that we have 42 apples and 17 oranges. It's not more complicated than that.

### The workflow of making graphs with `ggplot2`

If you were to build a house, you would normally start with the foundation and the basement. Then you add the main walls and the roof. Once the house's raw structure is done and stable, you would move on to work on improving how it looks: painting the facade, putting in windows, maybe doing some landscaping. 

The workflow of making graphs with `ggplot2` is very similar: The first thing we do is to create an empty graph window (the "foundation"). Then we define the main components of the graph, which dataset and variables we want to use and what type of graph we want to create (bar graph, line graph, scatter plot,...). Once these parts are done, we work on the finer details (labels, colors,...).

### The foundation: Opening an empty graph window

Everything in `ggplot2` starts with the `ggplot()` command (or 'function'). This is the main and most important function --- everything else is built on and around it. Simply put, this function tells R that we want to make a graph and to open a graph window.

Run the code below to see what happens when you run the `ggplot()` function without any settings:
``` {r fruit0, exercise=T}
ggplot()

```

###

``` {r fruit0a, exercise=F,echo=T}
ggplot()

```

There is not much going on --- all you see is an empty window. But consider what all we have told R to do with this command is open a graph window. Period. Not more. So all R does is to open an empty graph window.

We have not yet defined the most important things: Which dataset we want to use and what variable goes on which axis of the graph.

Let's add the missing information.

###

We provide this missing information by adding further specifications into the parentheses of the `ggplot()` function. Take a look at the code, and then run it to see what happens. We'll go over the code afterwards: 
```{r fruit_s1, exercise=T}
ggplot(data = fruit, aes(x = type, y = number))

```

###

```{r fruit_s1a, exercise=F,echo=T}
ggplot(data = fruit, 
       mapping = aes(x = type, y = number))

```

You may already guess what is going on here, but let's take a quick look together at what the code chunk did:

* As before, `ggplot()` tells R that we want to create a graph with ggplot.
* Within the `ggplot()`-command, we first specify that we want to use the fruit-dataset with `data=fruit`.
* The next part starting with `mapping` defines the so-called *aesthetic mapping*: Which variables we want to use, and how they should be placed into the graph. Here, `aes(...)` specifies the main dimensions (or _aesthetics_) of our graph. You can probably guess that we told R to put the type-variable on the x-axis at the bottom and the number-variable on the vertical y-axis.

Now we have laid the foundation for our graph! Let's add some content.

### Adding content

We add the main content on top of our foundation --- as an *additional layer*. As before, take a look at the code and then run it to see what it does. We'll then discuss what has happened:

```{r fruit2, exercise=T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col()

```

###

```{r fruit2a, exercise=F,echo=T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col()

```

Now we have a real graph! So, what happened? Note first that we have added an extra layer to our foundation with the `+` symbol. In `ggplot`, you always add layers with the `+` symbol.

We have specified this new layer as `geom_col()`, and obviously it has added bars --- or _columns_ --- to our graph. But what does this 'geom'-thing mean?

In `ggplot`, the core contents of a graph --- the bars in a bar graph, the dots in a scatterplot, the lines in a line graph --- are referred to as *geometric objects* or *geoms*'*.

For a simple bar graph, we use `geom_col()` --- i.e. 'geometric object: column'. We could have also used `geom_bar()` --- this is the more sophisticated version of `geom_col()` that offers more possibilities to tweak and customize the output. We'll use this one later.

There are also other *geometric objects* we could use:

* For a line graph, we would use `geom_line()`
* For a scatter plot, we would use `geom_point` (because a scatter plot shows, in essence, points or dots scattered over the graph)
* ...and many, many more...

This `geoms`-thing sounds perhaps a bit more complicated than necessary, but it does makes sense --- a bar, a dot or a line are all geometric objects! 

### Polishing the graph

Now that we have a first complete version of our graph, let's make it a bit prettier. The first thing we need to improve is the labels for the axes. As of now, it uses the 'raw' variable names. Before we can actually include this graph in a report, thesis, article, etc., we need to provide nicer-looking labels.

We do this again by adding new layers. You know the drill:
```{r fruit3, exercise=T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces")
```

###

```{r fruit3a, exercise=F, echo=T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces")
```

It should not be difficult to see what happened here. We have added two new layers to our graph:

* `xlab()` specifies the label of our x-axis
* `ylab` specifies the label of our y-axis

*Important*: Because axis labels are _text_, we need to put it in quotation marks!

###

As a last step, we change the overall look of the graph. The gray background is not ideal because it can sometimes make it more difficult to see smaller details in a graph and it also uses up more ink if we print it.

Typically, we want a sleek and simple black-and-white background. We can change the overall look -- or _theme_ -- of the graph as follows:
```{r fruit4, exercise=T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```

###

```{r fruit4a, exercise=F, echo=T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```

We have now changed the graph's theme by adding another new layer. Basically, the new layer `theme_bw()` overwrites the default gray theme and makes the graph black-and-white. There are a few other themes we could also have chosen: https://ggplot2.tidyverse.org/reference/ggtheme.html   

So, graph done! Not bad for a first go. I hope you also see that all this coding-stuff is less scary than it might seem.

Are you feeling ready to work on a real dataset? Then click 'continue'!

## Using a real dataset {data-progressive=TRUE}

In a real-life scenario, you would normally work with much larger datasets than the tiny fruit-dataset we used in the previous part. But the workflow to create graphs with ggplot stays generally the same: We set the foundation, we add the main contents, and then we polish.

In this second part of the tutorial, we will start working with real data --- specifically, we will use a part of the *Comparative Political Data Set* (https://www.cpds-data.org/). This dataset, which is widely used by comparative sociologists and political scientists in their research, includes information on a range of political and socio-economic variables for the economically advanced democracies since the 1960s --- for example which parties are in government, how high the unemployment rate is, and how many workers are organized in trade unions. It is also freely available --- you can use it, if you want!

The real dataset is quite large, but I have created a smaller subset of the dataset that includes only a few variables and covers only one year, 2015.

You can see this smaller dataset, which we will be working with now, below:
```{r cpds_tab, exercise=F, echo=F}
print.data.frame(cpds)
```

You can see that the dataset includes the following information:

* `country`: The name of the country for all 25 countries
* `iso`: The ISO 3-digit abbreviation for the country name
* `year`: The year, 2015
* `womenpar`: The share of female members of parliament
* `childcare`: How much each country spends on childcare, as a percentage of the size of its economy (its gross domestic product, or GDP)
* `unemp`: The country's unemployment rate
* `emprot_reg`: This is an index that measures how difficult it is for employers to dismiss ('fire') workers with a regular employment contract. This 'Employment Protection Index' has been created by the OECD. Higher scores mean it is more difficult to fire workers.

Notice also that there are some `NA`s in our dataset. These are observations where we have missing information ('Not Available'). In real-life research, you will see these very often!

### Which countries have more women in parliament?

Since we have already learned how to make a bar graph, let us use that skill to graph some real-world data. One interesting question we can look at with the CPDS data is which countries are better at including women in political processes. 

So, let's create the graph! I have created the *first piece* of the code necessary to create this graph for you: 

* You see that the dataset is saved as `cpds` (short for Comparative Political Dataset). 
* The `country`-variable is placed on the x-axis
* The `womenpar`-variable goes on the y-axis

Now you still need to add the correct 'geom' for a bar graph: `geom_col()`. Go ahead, add it to the code below --- and don't forget the `+`-symbol at the end of the first line!

```{r femrep1, exercise=T}
ggplot(data=cpds, aes(x=country,y=womenpar))

```

###

```{r femrep1a, exercise=F, echo=T}
ggplot(data=cpds, aes(x=country,y=womenpar)) +
  geom_col()

```

We have a graph -- but it certainly doesn't look very nice yet. Not only do we not have labelled axes and a clean theme, but we cannot even read the country names.

Let's work on the last problem.

###

One solution to the problem of overlapping country names is simply to flip the graph and make it horizontal so the bars go sideways instead of up. We can do this with `coord_flip()`. Can you add `coord_flip()` to the code below?

```{r femrep2, exercise=T}
ggplot(data=cpds, aes(x=country,y=womenpar)) +
  geom_col()

```

###

This is how the result should look like:
```{r femrep2a, exercise=F, echo=T}
ggplot(data=cpds, aes(x=country,y=womenpar)) +
  geom_col() +
  coord_flip()
```

Much better, isn't it?

###

Let's also add axis labels and adapt the theme. Can you add axis labels (with `xlab()` & `ylab()`) to the graph and adjust the theme (with `theme_bw()`)?
```{r femrep3, exercise=T}
ggplot(data=cpds, aes(x=country,y=womenpar)) +
  geom_col() +
  coord_flip()
```

###

This would be a solution:
```{r femrep3a, exercise=F, echo=T}
ggplot(data=cpds, aes(x=country,y=womenpar)) +
  geom_col() +
  coord_flip() +
  ylab("Share of women in parliament (%)") +
  xlab("Country") +
  theme_bw()
```

*Notice* that we still specify our axis labels as if the graph was not flipped!

### Sorting the bars

In the graph as it is now, the countries are still ordered alphabetically: Australia, Austria, Belgium, and so on. This is not bad, but it would be nicer if we could order everything by the share of women in parliament. That way we could easier see which countries do best and which do worst and representing women.

In 'ggplot', we can do that in the main `aes()`-command (the first line!), by specifying that we want the x-variable ordered by some other variable. You can see below how you do that. 

Take a look at the code, then run it to see the result:
```{r femrep4, exercise=T}
ggplot(data=cpds, aes(x=reorder(country, womenpar),y=womenpar)) +
  geom_col() +
  coord_flip() +
  ylab("Share of women in parliament (%)") +
  xlab("Country") +
  theme_bw()
```

###

```{r femrep4a, exercise=F, echo=T}
ggplot(data=cpds, aes(x=reorder(country, womenpar),y=womenpar)) +
  geom_col() +
  coord_flip() +
  ylab("Share of women in parliament (%)") +
  xlab("Country") +
  theme_bw()
```

So, what have we done? Note that we have added a `reorder()`-command to the main `aes()`-command. Within the `reorder()`-command, we first name the `country`-variable and then, after a comma, the `womenpar`-variable.

This tells R the following:

1. Take the `country`-variable
2. Order it by `womenpar`
3. Put the result on the x-axis

###

Note that we could also have ordered the order way around by adding a `-` like so: `reorder(country, -womenpar)`. Do you want to add the `-` and run the code:
```{r femrep4b, exercise=T}
ggplot(data=cpds, aes(x=reorder(country, womenpar),y=womenpar)) +
  geom_col() +
  coord_flip() +
  ylab("Share of women in parliament (%)") +
  xlab("Country") +
  theme_bw()
```

### Women's political power and the welfare state

We have learned that countries differ in how well its female citizens are represented in parliament. The next question is: Do more women in parliament have effects on policy-making?

Let's say we are interested in whether the share of female members of parliament in a country is related to the amount that country spends on childcare. Influential theories say that women should have a strong interest in well-developed childcare services [(Iversen & Rosenbluth 2010)](https://yalebooks.yale.edu/book/9780300171341/women-work-and-politics), and it is  possible that a larger share of female parliamentarians can push more forcefully for increased spending on these services.

To find out, we need to relate the `womenpar`-variable to the `childcare`-variable. The way we do this is with a _scatterplot_. A scatterplot is ideal if we want to show a relationship between two continuous (or metric, numerical) variables --- and, as we have seen, both of the variables are indeed continuous.

Below, you find again the *first piece* of the code necessary to create a scatterplot of these two variables. Run it to see what happens:

```{r cpds1, exercise=T}
ggplot(data = cpds, aes(x=womenpar,y=childcare))

```

You probably remember: This code only creates the foundation for our graph. We still need to add the core element, the _geometric object_, or 'geom'. Let's go do that!

###

Remember the correct 'geom' for scatterplots?
```{r q_geom, echo=F}
question("Pick the correct 'geom' for a scatterplot",
  answer("geom_bar()"),
  answer("geom_line()"),
  answer("geom_point()", correct = T),
  answer("geom_text()"),
  random_answer_order = TRUE,
  allow_retry = TRUE
)
```

###

So, we have to add a `geom_point()` to the existing code. You probably remember that you add extra layers with a `+`. Why don't you try this yourself: 

1. Add a `+` after the existing `ggplot()`-command;
2. On a new line, `geom_point()`. 

Then run the code.

```{r cpds2, exercise=T, warning=FALSE}
ggplot(data = cpds, aes(x=womenpar,y=childcare))
```

(Did you notice the auto-complete suggestions? Neat, right?)

###

Your code and graph should now look like this (if not, just scroll back up and klick 'start over'):
```{r cpds2a, exercise=F, echo=T, warning=FALSE}
ggplot(data = cpds, aes(x=womenpar,y=childcare)) +
geom_point()
```

The graph doesn't look too bad, and we can already learn something: There is indeed a positive relationship between the share of women in parliament and spending on childcare services.

But, as before, the graph still looks a bit unpolished without proper axis labels and with the dull gray background. So let's go and polish this graph like we did with the previous one!

###

Do you remember how to add axis labels with `xlab()` and `ylab()`? Why don't you give it a try and add these to the graph --- don't forget to always use the `+` symbol to add new layers!

```{r cpds3, exercise=T, warning=FALSE}
ggplot(data = cpds, aes(x=womenpar,y=childcare)) +
geom_point()
```

###

Your code and graph should now look like this (your axis labels might be different, but that is fine for now.):

```{r cpds3a, exercise=F,echo=T, warning=FALSE}
ggplot(data = cpds, aes(x=womenpar,y=childcare)) +
geom_point() +
xlab("Share of women in parliament (%)") +
ylab("Spending on childcare services (% GDP)")
```

Next we'll adapt the overall theme of the graph.

###

Can you change the graph theme to black-and-white (remember: `theme_bw()`):

```{r cpds4, exercise=T, warning=FALSE}
ggplot(data = cpds, aes(x=womenpar,y=childcare)) +
geom_point() +
xlab("Share of women in parliament (%)") +
ylab("Spending on childcare services (% GDP)")
```

###

Here is again how the result should look like now:

```{r cpds4a, exercise=F,echo=T, warning=FALSE}
ggplot(data = cpds, aes(x=womenpar,y=childcare)) +
geom_point() +
xlab("Share of women in parliament (%)") +
ylab("Spending on childcare services (% GDP)") +
theme_bw()
```

### Using text labels instead of dots

Technically, we have a quite neat-looking graph now. But we can still make it more informative. As it is now, the individual observations are only shown as black points. If we were working with survey data, where all observations are anonymous persons, this would be fine. But in this case, we know that these are all different countries --- and we know, of course, what these countries are called!

With 'ggplot', we can print the countries' names or the ISO abbreviations instead of the dots. Let's get going!

###

To use text labels, we need to use a different 'geom': `geom_text()`. So, to change the graph, we replace `geom_point()` with `geom_text()` --- but then there is some more!

To get started, take a look at the code I have prepared, run it, and then let's talk about what the code does:

```{r cpds6, exercise=T, warning=FALSE}
ggplot(data = cpds, aes(x=womenpar,y=childcare)) +
geom_text(aes(label=iso)) +
xlab("Share of women in parliament (%)") +
ylab("Spending on childcare services (% GDP)") +
theme_bw()
```

###

```{r cpds6a, exercise=F, echo=T, warning=FALSE}
ggplot(data = cpds, aes(x=womenpar,y=childcare)) +
geom_text(aes(label=iso)) +
xlab("Share of women in parliament (%)") +
ylab("Spending on childcare services (% GDP)") +
theme_bw()
```

Now, what has changed? Obviously, we have replaced `geom_point()` with `geom_text()` --- but we also have added some more information within `geom_text()`: there is new code to specify some central aesthetics (`aes()`) of the graph.

Let me explain: `geom_point()` is happy with knowing only which variables go on the x- and y-axes, and it then only puts a black dot for each observation at the appropriate point. 

`geom_text()`, on the other hand, is a bit more demanding: Like `geom_point()`, it needs to know the x- and y-variables, but it also needs to know how exactly each data point should be labeled. And we provide this information with an additional piece of _aesthetic_ (`aes()`) information. 

In our case, we want the graph to show the ISO abbreviations of country names, which are saved in the `iso` variable in our dataset. In other words, we want to 'map' the country abbreviations onto the graph. Therefore, we specify `aes(label=iso)`.

In this case, we added this extra `aes()` within `geom_text()` to make clear that this is specific to this 'geom'. We could, however, also have included it into the main `aes()`-function like this:
```{r cpds6b, exercise=T, warning=FALSE}
ggplot(data = cpds, aes(x=womenpar,y=childcare,label=iso)) +
geom_text() +
xlab("Share of women in parliament (%)") +
ylab("Spending on childcare services (% GDP)") +
theme_bw()
```

Run the code to see that it creates the same result.

All clear (more or less)? Well, great job! You just learned how to create a basic statistical graph in ggplot! Continue on to the next lesson to learn how to do this entirely by yourself.


## Doing it all yourself {data-progressive=TRUE}

### Prep work - Loading packages and data

If you were doing all this by yourself, the first step would be to load the ggplot2 package. Remember: R itself only offers some basic functionality and we use packages for 
more advanced things (or, often, to make our lives a bit easier...). Packages such as ggplot2 need to be installed and loaded. Below you see the code to install the ggplot2 package from the internet, fully automatically:

```{r eval=F,echo=T}
install.packages("ggplot2")
```

If you like, you can run this code in RStudio on your computer later. For now, we have installed the package and will now load it so that we can use it:
```{r eval=F,echo=T}
library(ggplot2)
```

The library-function is used to load packages.

We have also already loaded our CPDS-dataset. Let me just print it out again so you remember what it contains:
```{r cpds_tab2, exercise=F, echo=F}
print.data.frame(cpds)
```

* `country`: The name of the country for all 25 countries
* `iso`: The ISO 3-digit abbreviation for the country name
* `year`: The year, 2015
* `womenpar`: The share of female members of parliament
* `childcare`: How much each country spends on childcare, as a percentage of the size of its economy (its gross domestic product, or GDP)
* `unemp`: The country's unemployment rate
* `emprot_reg`: The index that measures how difficult it is for employers to dismiss ('fire') workers with a regular employment contract. Higher scores mean it is more difficult to fire workers.

Ready to move on?

### Are 'neoliberals' right?

Many conservative or liberal politicians (and many academics) argue that strict labor market regulation --- for instance making it difficult for employers to dismiss workers --- creates higher unemployment [(e.g. Siebert 1997)](https://www.aeaweb.org/articles?id=10.1257/jep.11.3.37).

With the CPDS data, we can find out if that is true! But now you have to do it yourself!

I have prepared a piece of code that you can start with --- *but you have to complete it for it to work!*

```{r neo1, exercise=T, warning=F}
ggplot(data=cpds,aes(x=,y=))
```
You need to:

1. Define which variables go on which axis (the relevant variables are called `unemp` & `emprot_reg`)
2. Add `geom_point()` to draw the scatterplot
3. Alternatively: Use ISO labels with `geom_text()`
4. Add informative axis labels with `xlab()` & `ylab()`
5. Change the overall theme with `theme_bw()` --- or another one, if you have a different preference!

### My solution

Here is how I would have done it:
```{r neo2, exercise=F, echo=T, warning=F}
ggplot(data=cpds,aes(x=emprot_reg,y=unemp)) +
  geom_text(aes(label=iso)) +
  ylab("Unemployment rate (%)") +
  xlab("Strictness of dismissal rules") +
  theme_bw()
```

What do we learn? There is indeed a positive relationship: Countries with stricter labor market rules like Greece or Spain have higher unemployment rates than countries with lenient rules like the United States. But: The relationship is not particularly strong!

Anyways, graph number 3 done --- and this time (mostly) yourself! 

## De-bugging exercises {data-progressive=TRUE}

So far, you have been given code examples that worked --- in the sense that they did not cause an error. If you were to do your own data analysis, you would of course eventually run into a problem and/or receive error messages. The exercises in this chapter prepare you for that.

Specifically, this chapter should help you achieve the following:

1) You get more practice working with the code examples from earlier, which will reinforce the skills you have acquired so far
2) You get so see the most common causes of error messages ('bugs') --- and that they are usually easily fixable and less scary than they might seem at first
3) You get used to dealing with problems when coding, and thus become a more confident user of R

###

### Exercise 1

Below you see a set of code for a bar graph that does not work. Click 'run code' to see the error message (scary, right?). Then try to fix it. There is a hint, if you need it.
```{r debug1, exercise = T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col( +
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```
<div id="debug1-hint">
**Hint:** Check that all parentheses -- ( ) -- are closed.
</div>

### 

You only needed to close the parenthesis for `geom_col`:
```{r debug1a, eval=T, echo=T, exercise=F}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```


### Exercise 2

Here is one more code chunk that throws an error. Can you fix this one?
```{r debug2, exercise = T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col() +
    xlab(Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```
<div id="debug2-hint">
**Hint:** Check that all text is enclosed in quotation marks ("").
</div>

### 

The solution was to add an opening quotation mark before "Type of fruit":
```{r debug2a, eval=T, echo=T, exercise=F}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```

### Exercise 3

And one more, with another very common bug:
```{r debug3, exercise = T}
ggplot(data = fruit, aes(x = type y = number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```
<div id="debug3-hint">
**Hint:** In R, all elements of a function (what is inside a function's parentheses, e.g. `aes()`) need to be separated by commas. Carefully go over the first line -- can you see two elements of a function that are *not* separated by a comma?
</div>

###

The trick was to place a comma between the x- and y-elements of the `aes()` function:
```{r debug3a, exercise = F, eval=T, echo=T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```

### Exercise 4

This one might be a bit tricky to spot:
```{r debug4, exercise = T}
ggplot(data = fruits, aes(x = typ, y = number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```
<div id="debug4-hint">
**Hint:** The dataset is called `fruit`, and the variable for the type of fruit is called `type`.
</div>

###

The solution: Remove the s in `fruits` and change `typ` to `type`:
```{r debug4a, exercise = F, eval=T, echo=T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```

### Exercise 5

If you paid attention earlier, you should be able to spot this one easily:
```{r debug5, exercise = T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col() 
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```
<div id="debug5-hint">
**Hint:** You add extra layers to a graph with a `+` symbol.
</div>

###

The `+` symbol behind `geom_col()` was missing!
```{r debug5a, exercise = F, eval=T, echo=T}
ggplot(data = fruit, aes(x = type, y = number)) +
    geom_col() +
    xlab("Type of fruit") +
    ylab("Number of pieces") +
    theme_bw()
```


### Summary

These exercises introduced you to the (in my experience) most common sources of error messages:

* Parentheses that were opened "("  but not also closed ")". This is confusing the computer because it is desperately looking for the end of the function that is lacking the closing parenthesis.
* Text that is not enclosed in quotation marks. Computers can only read pre-defined code --- their native language. They do not understand human language.
* Missing commas. This can happen easily when one writes code in a hurry. R needs you to be very explicit about how the different parts of a function should be separated.
* Misspelled dataset- or variable names. This one is particularly devious: It is actually good if it results in an error because you referred to a non-existing dataset (`fruits`) or variable (`typ`). It is much more problematic if there actually _is_ a dataset called `fruits` because then the code runs error-free but might show you the wrong results!

But you should also have seen that these 'bugs' are usually easy and fast to fix. You may at some point run into some problems that are more tricky --- but this happens very rarely. Mostly, it is stray commas, missing symbols, quotation marks and parentheses.

*Now you should be ready to do it all yourself, without 'training wheels'! Bonne chance!*
